Species persistence under climate and fishing

### 2014-02-03 Writing/MS.tex
added Catherine and Will to acknowledgments
last of malins comments

We would like to know the rate of environmental shift and the harvesting rate such that as long as the 
environment moves more slowly or we harvest less severely than those parameters, then the population will be able to persist. We call these, respectively, the critical rate of environmental shift, $c^*$, and the critical harvesting rate, $h^*$. 
We found these rates by finding the parameters that make the trivial pulse unstable.   In our analyses, we used a  separable kernel, the Gaussian kernel given by

@@

\[ k(x-y)=\frac{1}{2}be^{-b|x-y|}.\]

\noindent Second, we examined harvesting rules more complex than harvesting a constant proportion of the population. Whereas population persistence in the analytical model does not depend on the functional form of recruitment $f$, to perform simulations we must 
specify a recruitment function.  Again, we chose to use a Beverton-Holt function.  In the first generation, we seeded the world with 50 individuals at a single point, as in \citep{ZhouKot2011}. We first ran through 150 generations in order for the population to reach equilibrium without harvesting or climate shift.  We then added harvesting pressure, allowed the population to again reach equilibrium, and finally added climate change by moving the viable patch.  Equilibrium biomass is calculated as the mean biomass of 300 time steps once the difference in biomass between time step $t$ and $t+1$ was no greater than $0.1$.

@@

Knowing whether two disturbances are synergistic in their impacts for a given population is important for management. The co-occurrence of climate change-driven range shifts and fishing mean that these is the potential for synergistic interactions, which have been largely unexamined.  Previous modeling work examining examining how harvest and range shift intersect has lacked the generality to examine how these two disturbances are likely to jointly affect population persistence. Many modeling studies have focused on how fish populations are likely to change in the future under a variety of fishing and climate scenarios, but the degree of detail and case-specificity makes it difficult to draw general conclusions. Here we have built a general model to examine how climate and harvesting interact to affect species persistence by incorporating dispersal and reproduction.

@@

While we did find synergy between the stressors, the drop in biomass caused by both stressors was never very much higher than the null prediction, i.e. synergistic effects were quite small.  However, the fact that synergy is highest in those populations whose persistence is most tenuous is worrisome from a conservation perspective.  Additionally, the fact that we found additive synergy means that harvesting levels or rate of environment shift that are sustainable individually together can drive a population to extinction.  Synergy between harvesting and climate changes, e.g. warming temperature or a more variable climate, has been identified in experiments \citep{Moraetal2007}, in the demography and size of specific populations \citep{Planque:2010uq}, and at the ecosystem level \citep{Kirby:2009fk}.  Our results  agree with these empirical findings and it is therefore important to take the effects of both stressors into consideration when designing conservation and management strategies.
% the below needs some help, Eleanor can you flesh out with a better discussion of how our results compare to Mora, Planque and Kirby?

@@

\section*{ Acknowledgements}
We'd like to thank our reviewers for providing constructive comments.

@@

\begin{table}
\begin{tabular}{@{}lllllll@{}}
  Variable & Definition
\\\cmidrule{1-1} \cmidrule{2-2}   
$n_t(x)$ & density of fish at position $x$ at time $t$
\\ $n^*(\overline{x})$ & density of fish at equilibrium at position $\overline{x}$ relative to the patch 
\\ $k(x-y)$ & dispersal kernel, probability of larva traveling from position $y$ to position $x$
\\ $\langle d \rangle $ & expected distance traveled by larva
\\ $f(n)$ & recruitment function, number of offspring produced by a population of size $n$
\\ $R_0$ & intrinsic growth rate, $R_0=f'(0)$
\\ $h$ & proportion of adults harvested
\\ $L$ & patch length
\\ $c$ & rate of environmental shift
\end{tabular}
\caption{}
\label{variables}
\end{table}

\section{Discussion}
We extended a spatial model to study a fish population subject to both the climate-driven range shifts and harvesting. Previous work has highlighted the importance of reproduction and dispersal to the vulnerability of fish populations to climate change \citep[\citet{Fordhametal2013}]{Hastingsetal2005} and our model explicitly includes these processes.  In particular, we derived expressions for critical harvesting rates and critical rates of environmental shift for a separable Gaussian kernel and found these critical rates using numerical simulations of a Laplace kernel.  For each kernel we studied, we found that the higher the growth rate and the better the mean dispersal distance matches the rate of environmental shift, the better a population can adjust to harvest and climate change.  More interestingly, we found a negative relationship between the critical harvesting rate and the rate of environmental shift.  That is, the more quickly the environment shifts the less harvesting it takes to drive the population extinct.  The curved line separating parameters that will allow the population to persist from those that won't is on indication of an interaction between the stressors.

While we did find synergy between the stressors, the drop in biomass caused by both stressors was never very much higher than the null prediction, i.e. synergistic effects were quite small.  However, the fact that synergy is highest in those populations whose persistence is most tenuous is worrisome from a conservation perspective.  Additionally, the fact that we found additive synergy means that harvesting levels or rate of environment shift that are sustainable individually together can drive a population to extinction.  Synergy between harvesting and climate changes, e.g. warming temperature or a more variable climate, has been identified in experiments \citep{Moraetal2007}, in the demography and size of specific populations \citep{Planque:2010uq}, and at the ecosystem level \citep{Kirby:2009fk}.  Our results  agree with these empirical findings and it is therefore important to take the effects of both stressors into consideration when designing conservation and management strategies.

The two management strategies we modeled -- harvesting thresholds and MPAs -- both increased the population's biomass at equilibrium and improved its ability to persist.  Protected areas have been advanced as a way to help organisms keep pace with range shifts, as well as to ameliorate anthropogenic disturbances like harvesting and habitat fragmentation\citep{Lawleretal2010, Hannahetal2007,Botsfordetal2001, Gaylordetal2005, HastingsBotsford2003,Thomasetal2012}.  Our results, that spatial management increased the maximum harvesting rate and maximum climate velocity at which the population could survive, corroborate these previous findings.  The management strategies also affect how the two stressors interact with each other.  Specifically, harvesting thresholds can alleviate interactions between the two stressors.   While the management strategies only change harvesting practices and do not directly address affect climate change, understanding how they ameliorate synergistic affects between harvesting and range shifts will help to better implement and place protected areas.

Finally, our results suggest that particular combinations of harvesting and rate of environmental shift will affect some species more than others. Indeed, \citet{Perryetal2005} found that fish that shifted in response to warming in North Sea had faster life histories than non shifting species (smaller body sizes, faster maturation, smaller sizes at maturity). Using a simple mechanistic model like the one we presented can incorporate these species interactions and multiple disturbances to understand whether specific life histories are likely to be selected over others as harvesting and/or range shifts increase.

>>>>

We would like to know the rate of environmental shift and the harvesting rate such that as long as the 
environment moves more slowly or we harvest less severely than those parameters, then the population will be able to persist. We call these, respectively, the critical rate of environmental shift, $c^*$, and the critical harvesting rate, $h^*$. 
We found these rates by finding the parameters that make the trivial pulse unstable.   In our analyses, as in \citep{Latore:1998fk}, we used the separable Gaussian kernel given by

@@

\[ k(x-y)=\frac{1}{2}be^{-b|x-y|},\]

\noindent a commonly used model of larval dispersal \citep{Pinsky:2011fk}.  Second, we examined harvesting rules more complex than harvesting a constant proportion of the population. Whereas population persistence in the analytical model does not depend on the functional form of recruitment $f$, to perform simulations we must 
specify a recruitment function.  Again, we chose to use a Beverton-Holt function.  In the first generation, we seeded the world with 50 individuals at a single point, as in \citep{ZhouKot2011}. We first ran through 150 generations in order for the population to reach equilibrium without harvesting or climate shift.  We then added harvesting pressure, allowed the population to again reach equilibrium, and finally added climate change by moving the viable patch.  Equilibrium biomass is calculated as the mean biomass of 300 time steps once the difference in biomass between time step $t$ and $t+1$ was no greater than $0.1$.

@@

Knowing whether two disturbances are synergistic in their impacts for a given population is important for management. The co-occurrence of climate change-driven range shifts and fishing mean that there is the potential for synergistic interactions, which have been largely unexamined.  Previous work examining  how harvest and range shift interact has lacked the generality to examine how these two disturbances are likely to jointly affect population persistence. Many modeling studies have focused on how fish populations are likely to change in the future under a variety of fishing and climate scenarios, but the degree of detail and case-specificity makes it difficult to draw general conclusions. Here we have built a general model to examine how climate and harvesting interact to affect species persistence by incorporating dispersal and reproduction.

@@

While we did find synergy between the stressors, the drop in biomass caused by both stressors was never very much higher than the null prediction, i.e. synergistic effects were quite small.  However, the fact that synergy is highest in those populations whose persistence is most tenuous is worrisome from a conservation perspective.  Additionally, the fact that we found additive synergy means that harvesting levels or rate of environment shift that are sustainable individually together can drive a population to extinction.  Synergy between harvesting and climate changes has been identified in experimental populations \citep{Moraetal2007}, in specific populations \citep{Planque:2010uq}, and at the ecosystem level \citep{Kirby:2009fk,Planque:2010uq}.  However, these results are not directly comparable to ours because they focus on different aspects of climate change, e.g. warming temperature \citep{Moraetal2007,Kirby:2009fk} or a more variable climate \citep{Planque:2010uq}.  Additionally, in the experimental populations, synergy was identified between warming and harvesting but not between habitat fragmentation \citep{Moraetal2007}.  Nevertheless, our results generally agree with these empirical findings and it is therefore important to take the effects of both stressors into consideration when designing conservation and management strategies.

@@

\section*{ Acknowledgements}
We thank Catherine Offord and Will Scott for their contributions to an earlier version of this manuscript and our reviewers for providing constructive comments.

@@

\begin{table}
\caption{Table of variables used in the text}
\begin{tabular}{@{}lllllll@{}}
  Variable & Definition
\\\cmidrule{1-1} \cmidrule{2-2}   
$n_t(x)$ & density of fish at position $x$ at time $t$
\\ $n^*(\overline{x})$ & density of fish at equilibrium at position $\overline{x}$ relative to the patch 
\\ $k(x-y)$ & dispersal kernel, probability of larva traveling from position $y$ to position $x$
\\ $\langle d \rangle $ & expected distance traveled by larva
\\ $f(n)$ & recruitment function, number of offspring produced by a population of size $n$
\\ $R_0$ & intrinsic growth rate, $R_0=f'(0)$
\\ $h$ & proportion of adults harvested
\\ $L$ & patch length
\\ $c$ & rate of environmental shift
\end{tabular}
\label{variables}
\end{table}

%\section{Discussion}
%We extended a spatial model to study a fish population subject to both the climate-driven range shifts and harvesting. Previous work has highlighted the importance of reproduction and dispersal to the vulnerability of fish populations to climate change \citep[\citet{Fordhametal2013}]{Hastingsetal2005} and our model explicitly includes these processes.  In particular, we derived expressions for critical harvesting rates and critical rates of environmental shift for a separable Gaussian kernel and found these critical rates using numerical simulations of a Laplace kernel.  For each kernel we studied, we found that the higher the growth rate and the better the mean dispersal distance matches the rate of environmental shift, the better a population can adjust to harvest and climate change.  More interestingly, we found a negative relationship between the critical harvesting rate and the rate of environmental shift.  That is, the more quickly the environment shifts the less harvesting it takes to drive the population extinct.  The curved line separating parameters that will allow the population to persist from those that won't is on indication of an interaction between the stressors.
%
%
%To quantify the interaction between the stressors, we measured the synergy between their effects on population biomass.  We found positive synergy between the stressors and that the synergy is greatest in the region of parameter space where the equilibrium biomass is smallest.  We found similar results from the analytically derived biomass and the simulation derived biomass. This indicates that this result is robust to changes in the dispersal kernel.  We chose to measure the effect of each stressor by the absolute drop in biomass caused by the stressor, and we used the sum of the individual effects for our null prediction of the effect of both stressors, as in \citep{Crainetal2008, DarlingCote2008,Nyeetal2013}.  We could also have measured the effect by the percentage drop caused by the stressor(s) and used a multiplicative null prediction for the effect of both stressors.  In general, measuring synergy against an additive null prediction is more conservative than measuring synergy multiplicatively: the presence of additive synergy implies multiplicative synergy, but not vice versa \citep{Crainetal2008, Foltetal1999}.  Since we found small levels of positive additive synergy between the two stressors, other measures of synergy might show even higher levels of interaction. 
%
%While we did find synergy between the stressors, the drop in biomass caused by both stressors was never very much higher than the null prediction, i.e. synergistic effects were quite small.  However, the fact that synergy is highest in those populations whose persistence is most tenuous is worrisome from a conservation perspective.  Additionally, the fact that we found additive synergy means that harvesting levels or rate of environment shift that are sustainable individually together can drive a population to extinction.  Synergy between harvesting and climate changes, e.g. warming temperature or a more variable climate, has been identified in experiments \citep{Moraetal2007}, in the demography and size of specific populations \citep{Planque:2010uq}, and at the ecosystem level \citep{Kirby:2009fk}.  Our results  agree with these empirical findings and it is therefore important to take the effects of both stressors into consideration when designing conservation and management strategies.
%
%The two management strategies we modeled -- harvesting thresholds and MPAs -- both increased the population's biomass at equilibrium and improved its ability to persist.  Protected areas have been advanced as a way to help organisms keep pace with range shifts, as well as to ameliorate anthropogenic disturbances like harvesting and habitat fragmentation\citep{Lawleretal2010, Hannahetal2007,Botsfordetal2001, Gaylordetal2005, HastingsBotsford2003,Thomasetal2012}.  Our results, that spatial management increased the maximum harvesting rate and maximum climate velocity at which the population could survive, corroborate these previous findings.  The management strategies also affect how the two stressors interact with each other.  Specifically, harvesting thresholds can alleviate interactions between the two stressors.   While the management strategies only change harvesting practices and do not directly address affect climate change, understanding how they ameliorate synergistic affects between harvesting and range shifts will help to better implement and place protected areas.
%
%The advantage of a simple model like ours is that it is general enough to be applied to a number of systems.  However, it  ignores many of the complexities present in marine fisheries. We do not include Allee effects, so that even if the population shrank to very low levels it was possible for it to persist over time. However, we found that qualitatively similar results about the interaction between climate and harvesting would hold for a model with a recruitment function with Allee effect. We also did not include age structure in our model. The effects of both harvesting and climate change may be different across different age classes; including this level of complexity is left for future work. Additionally, we did not include any mechanisms aside from larval dispersal by which the population could keep up with a shifting climate.  In addition to these species-specific extensions, this modeling framework could be extended to consider species interactions, especially predator-prey pairs.  By introducing a predatory species, we would be imposing yet another stressor on the focus species \citep{Lingetal2009, Gurevitchetal2000}and we are interested in measuring the interaction between the effects of this stressor and the two we consider here.
%
%Finally, our results suggest that particular combinations of harvesting and rate of environmental shift will affect some species more than others. Indeed, \citet{Perryetal2005} found that fish that shifted in response to warming in North Sea had faster life histories than non shifting species (smaller body sizes, faster maturation, smaller sizes at maturity). Using a simple mechanistic model like the one we presented can incorporate these species interactions and multiple disturbances to understand whether specific life histories are likely to be selected over others as harvesting and/or range shifts increase.
